State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-13 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
changed table1
changed the units in Table 1 - only one left to do is
made a large number of smallish changes. Please run a diff to check between versions and see what you thinkcd Dropbox/Projects/BrainIDE/cd Dropbox/Projects/BrainIDE/

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. We use the reduced model to estimate the parameters of the full neural field model. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\section{Introduction} This paper presents a theoretical framework for generating patient-specific macroscopic neural field mathematical models from data. There has been much interest in generating physiologically plausible neural field models to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is this scale at which pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested. Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been reported (to the best of the authors knowledge). Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development from this recent paper was to represent the spatiotemporal system as a standard state-space model, with the number of states independent of the number of observation locations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields.

Modelling the neural dynamics within this framework has a distinct advantage over the more standard multivariate auto-regressive (AR) models: that the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity.

In this paper, we demonstrate for the first time how intracortical connectivity can be inferred from data, based on a variant of the Wilson and Cowan neural field model~\cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and patient-specific data.

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a reduced finite dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} establishes conditions using spatial frequency analysis for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments.

\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC neural field model that was used to create our intracortical connectivity estimator. Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex patterns of activity such as Turing patterns, spirals, and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

\begin{figure}[th]\label{fig:2d_kernel}

@@

   	\caption{A mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$ used in this paper. The function $w(.)$ is rotationally symmetric (isotropic) about zero, hence a cross section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero with space.} 

@@

where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a spatial Gaussian process with mean zero and covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. This term is added to account for uncertainty in the model. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper. The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors, $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ and $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 

@@

Under the assumption the basis function decomposition is accurate, the observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into~\ref{ObservationEquation} giving

@@

Let the spectral representation of the dynamic field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. According to Shannon's sampling theorem the field must be spatially band-limited, for an accurate reconstruction using spatially discrete observations. Such a condition ensures that $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as -3~dB point) with $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y < \frac{1}{2\rho_y\nu_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects when reconstructing the hidden dynamic field $v_t(\mathbf{r})$ using the sampled observations, $\mathbf{y}_t$.

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, thus allowing the positioning of sensor in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate it using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors spectral characteristics. Using spatially extended sensors results in spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore band-limit the field to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

Although such sensors avoid errors due to aliasing, they attenuate the high spatial frequency variations in the observations. Therefore, any procedure applied to estimate the original field or the underlying connectivity structure from these band-limited observations has the potential to underestimate the high spatial frequency components of both the field and the connectivity kernel. Such a remark motivates the use of sensors with wider bandwidths (narrower in space). Nevertheless, this choice would require more sensors in order to satisfy Shannon's sampling theorem for a reasonable spatial region. Therefore, a compromise needs to be found between the bandwidth of the sensor response, the accuracy of the estimation results, the number of sensors used and thus the computational demands of estimation procedure when designed experiments.

Similar considerations need to be made regarding the representation of the dynamic field, $v_t(\mathbf{r})$, using the basis function decomposition. The minimum distance between adjacent basis functions must also satisfy Shannon's sampling theorem to capture \dean{should this be `represent' instead of `capture'?} the dominant spectral information of the field. Thus, the minimum distance between basis functions must satisfy 
\begin{equation}\label{eq:BasisFunctionSeparation}
	\Delta_{\phi} < \frac{1}{2\rho_{\phi}\nu_{cy}}
\end{equation}
where $\rho_{\phi}$ is an oversampling parameter to determine the basis function separation.

@@

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe the procedure for estimating the states, $\mathbf{x}_t$, and the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. At each iteration, the sequence of estimated state vectors is used to update the parameter set. The resulting parameters are then used when estimating the new state vector sequence for the next iteration. The procedure stops when the parameters converge. The algorithm is initialised using a bounded random state vector sequence that guarantees the initial estimated parameter set forms a stable kernel.

@@

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the URTSS, we can write
\begin{eqnarray}
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\
\end{eqnarray}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the parameters, $ \mathbf W$, can be estimated by
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}
% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

\begin{table*}\footnotesize

@@

\begin{tabular}{c|c|c}

@@

	$\zeta$ & 0.01 & ms$^{-1}$\\

@@

	\right]^{\top}$ & synaptic strength\\
	$\sigma_{\psi,1}$ & 1.8 & inner connectivity kernel width\\
	$\sigma_{\psi,2}$ & 2.4 & middle connectivity kernel width\\
	$\sigma_{\psi,3}$ & 6 & outer connectivity kernel width\\
	$\Sigma$ & & field disturbance covariance \\
	sensor parameters & & \\
	$\Delta$ & 0.5 & space step \\
	$T_s$ & 1 & ms time step \\ 

@@

\end{table*}

@@

\begin{figure}[th]

@@

\subfigure[][]{\label{fig:FieldFFT}\includegraphics[scale=1]{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics[scale=1]{./Graph/FFTEstimatedField.pdf}}\\

@@

The spatial frequency from the observations is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}$, of $1$, the minimum distance between basis functions was calculated to be $2.63$ to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). From equation~\ref{eq:WidthCutOffRelationship} the required basis function width to represent the dominant spatial frequencies was calculated to be $1$. The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$, which is sufficient to capture the significant dynamics according to our criteria. The basis function width, $\sigma_{\phi}^2$, was set above the calculated width to $2.5$, which reflected a design choice. This was a result of a trade-off that ensured that the lower spatial frequency components of the field were better represented at the expense of higher spatial frequency components. The spatial frequency response of the basis function can be seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure*} 

@@

		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics[scale=1]{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics[scale=1]{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics[scale=1]{./Graph/BasisChar.pdf}}

@@

\begin{figure*}

@@

\end{figure*}
\begin{figure}[th]
\includegraphics[scale=1]{./Graph/convergence.pdf}

@@

>>>>

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. We use the reduced model to estimate the parameters of the full neural field model \mike{not sure what this last sentence meant?}. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\section{Introduction}

Generating physiologically plausible neural field models are of great importance in order to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is this scale at which pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

We present an extension to the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been reported (to the best of the authors' knowledge). Our work builds on recent work which shows that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development here was to represent the spatiotemporal system as a standard state-space model, with the number of states independent of the number of observation locations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields.

Modelling the neural dynamics within this framework has a distinct advantage over the more standard multivariate auto-regressive (AR) models: the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity.

In this paper, we demonstrate for the first time how intracortical connectivity can be inferred from data, based on a variant of the WC neural field model~\cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and patient-specific data.

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a finite dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} establishes conditions using spatial frequency analysis for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments.

\section{Neural Field Model}\label{NeuralModelSection}

Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex patterns of activity such as Turing patterns, spirals, and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

\begin{figure}\label{fig:2d_kernel}

@@

   	\caption{A mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$ used in this paper. The function $w(.)$ is rotationally symmetric (isotropic) about zero, hence a cross section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero.} 

@@

where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a spatial Gaussian Process with mean zero and covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. This term is added to account for uncertainty in the model. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}
    \label{eq:ObservationEquation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors, $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ and $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

Under the assumption the basis function decomposition is accurate, the observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into~\ref{eq:ObservationEquation} giving

@@

Let the spectral representation of the post-synaptic membrane voltage field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. According to Shannon's sampling theorem the field must be spatially band-limited, for an accurate reconstruction using spatially discrete observations. Such a condition ensures that $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as -3~dB point) with $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y < \frac{1}{2\rho_y\nu_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects when reconstructing the hidden dynamic field $v_t(\mathbf{r})$ using the sampled observations, $\mathbf{y}_t$.

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, stopping us from positioning sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate it using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques~\cite{Issa2000}. Nevertheless, spectral aliasing can still be avoided by a proper choice of the spatial sampling distance given the sensors spectral characteristics. The spatial extent of the sensors results in spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors therefore can only sense the field up to a known bandwidth, denoted by $\nu_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, with $\nu_{cy}$ replacing $\nu_c$.

Although such sensors avoid errors due to aliasing, they attenuate the high spatial frequency variations in the observations. Therefore, any procedure applied to estimate the original field or the underlying connectivity structure from these band-limited observations has the potential to underestimate the high spatial frequency components of both the field and the connectivity kernel. Such a remark motivates the use of sensors with wider bandwidths (narrower in space). Nevertheless, this choice would require more sensors in order to satisfy Shannon's sampling theorem for a reasonable spatial region. Therefore, a compromise needs to be found between the bandwidth of the sensor response, the accuracy of the estimation results, the number of sensors used and the computational demands of the estimation procedure when designing experiments.

Similar considerations need to be made regarding the representation of the dynamic field, $v_t(\mathbf{r})$, using the basis function decomposition. Again applying Shannon's Sampling Theorem, the minimum distance between basis functions must satisfy 
\begin{equation}\label{eq:BasisFunctionSeparation}
	\Delta_{\phi} < \frac{1}{2\rho_{\phi}\nu_{cy}}
\end{equation}
where $\rho_{\phi}$ is an oversampling parameter to determine the basis function separation.

@@

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe the procedure for estimating the states, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. At each iteration, the sequence of estimated state vectors is used to update the parameter set. The resulting parameters are then used when estimating the new state vector sequence for the next iteration. The procedure stops when the parameters converge. The algorithm is initialised using a bounded random state vector sequence that guarantees the initial estimated parameter set forms a stable kernel.

@@

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the URTSS, we can write
\begin{eqnarray*}
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1  \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. 
\end{eqnarray*}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the parameters, $ \mathbf W$, can be estimated using
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}
% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

\begin{table}\footnotesize

@@

\begin{tabular}{ccl}

@@

	$\tau$ & 0.01 & ms$^{-1}$\\

@@

	\right]^{\top}$ & mV/spike\\
	$\sigma_{\psi_{1}}, \sigma_{\psi_{2}}, \sigma_{\psi_{3}}$ & 1.8, 2.4, 6 & mm\\
	$\sigma_{m},\Delta_{y}$&0.9, 1.5&mm\\
	$\Sigma_{\epsilon}$ &$0.1 \times I_{N} $& mm$^2$ \\
	$\sigma_{\gamma}$&1.3&mm\\
	$\Delta$ & 0.5 & mm \\
	$T_s$ & 1 & ms \\ 

@@

\end{table}

@@

@@

\subfigure[][]{\label{fig:FieldFFT}\includegraphics{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics{./Graph/FFTEstimatedField.pdf}}\\

@@

The spatial frequency from the observations is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}$, of $1$, the minimum distance between basis functions was calculated to be $2.63$ to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). From equation~\ref{eq:WidthCutOffRelationship} the required basis function width to represent the dominant spatial frequencies was calculated to be $1$. The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$, which is sufficient to capture the significant dynamics according to our criteria. The basis function width, $\sigma_{\phi}^2$, was set above the calculated width to $2.5$, which reflected a design choice. This was a result of a trade-off that ensured that the lower spatial frequency components of the field were better represented at the expense of higher spatial frequency components \mike{I don't know why there's a tradeoff between estimating low and high frequency components of the kernel, and I'm not sure I've been convinced by this section}. The spatial frequency response of the basis function can be seen in Figure~\ref{fig:BasisFunctionFreqResponse}.

@@

		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}

@@

@@

\includegraphics{./Graph/convergence.pdf}

@@

\section{Conclusion}
